Experiment 2 (of 4)

### 2016-08-07 experiments/e2/writeup/e2.Rmd
small edit
compiled
tidied up e2 considerably, switched to pdf output

---
title: "Experiment 2 (of 4)"
author: "Matt Green"
output: 
  html_document:
    toc: true
    <!--toc_float: true-->
    mathjax: default
    template: default
    code_folding: hide
    theme: default
---

```{r loadKnitr}
suppressPackageStartupMessages(library(knitr))
```

```{r loadLibs}
suppressPackageStartupMessages(library(MASS))
suppressPackageStartupMessages(library(xtable))
suppressPackageStartupMessages(library(lattice))
suppressPackageStartupMessages(library(latticeExtra))
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(library(beepr))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(lmerTest))
suppressPackageStartupMessages(library(LMERConvenienceFunctions)) # for perSubjectTrim
suppressPackageStartupMessages(library(languageR))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(plyr))
suppressPackageStartupMessages(library(rms))                      # for varclus
suppressPackageStartupMessages(library(languageR))
```

```{r CIfunctions}
## Summarizes data.
## Gives count, mean, standard deviation, standard error of the mean, and confidence interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
  require(plyr)
  # New version of length which can handle NA's: if na.rm==T, don't count them
  length2 <- function (x, na.rm=FALSE) {
    if (na.rm) sum(!is.na(x))
    else       length(x)
  # This is does the summary; it's not easy to understand...
  datac <- ddply(data, groupvars, .drop=.drop,
                 .fun= function(xx, col, na.rm) {
                   c( N    = length2(xx[,col], na.rm=na.rm),
                      mean = mean   (xx[,col], na.rm=na.rm),
                      sd   = sd     (xx[,col], na.rm=na.rm)
                 },
                 measurevar,
                 na.rm
  # Rename the "mean" column    
  datac <- rename(datac, c("mean"=measurevar))
  datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
  # Confidence interval multiplier for standard error
  # Calculate t-statistic for confidence interval: 
  # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
  ciMult <- qt(conf.interval/2 + .5, datac$N-1)
  datac$ci <- datac$se * ciMult
  return(datac)
## Summarizes data, handling within-subjects variables by removing inter-subject variability.
## It will still work if there are no within-S variables.
## Gives count, un-normed mean, normed mean (with same between-group mean),
##   standard deviation, standard error of the mean, and confidence interval.
## If there are within-subject variables, calculate adjusted values using method from Morey (2008).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   betweenvars: a vector containing names of columns that are between-subjects variables
##   withinvars: a vector containing names of columns that are within-subjects variables
##   idvar: the name of a column that identifies each subject (or matched subjects)
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySEwithin <- function(data=NULL, measurevar, betweenvars=NULL, withinvars=NULL,
                            idvar=NULL, na.rm=FALSE, conf.interval=.95, .drop=TRUE) {
  # Ensure that the betweenvars and withinvars are factors
  factorvars <- vapply(data[, c(betweenvars, withinvars), drop=FALSE],
                       FUN=is.factor, FUN.VALUE=logical(1))
  if (!all(factorvars)) {
    nonfactorvars <- names(factorvars)[!factorvars]
    message("Automatically converting the following non-factors to factors: ",
            paste(nonfactorvars, collapse = ", "))
    data[nonfactorvars] <- lapply(data[nonfactorvars], factor)
  # Get the means from the un-normed data
  datac <- summarySE(data, measurevar, groupvars=c(betweenvars, withinvars),
                     na.rm=na.rm, conf.interval=conf.interval, .drop=.drop)
  # Drop all the unused columns (these will be calculated with normed data)
  datac$sd <- NULL
  datac$se <- NULL
  datac$ci <- NULL
  # Norm each subject's data
  ndata <- normDataWithin(data, idvar, measurevar, betweenvars, na.rm, .drop=.drop)
  # This is the name of the new column
  measurevar_n <- paste(measurevar, "_norm", sep="")
  # Collapse the normed data - now we can treat between and within vars the same
  ndatac <- summarySE(ndata, measurevar_n, groupvars=c(betweenvars, withinvars),
                      na.rm=na.rm, conf.interval=conf.interval, .drop=.drop)
  # Apply correction from Morey (2008) to the standard error and confidence interval
  #  Get the product of the number of conditions of within-S variables
  nWithinGroups    <- prod(vapply(ndatac[,withinvars, drop=FALSE], FUN=nlevels,
                                  FUN.VALUE=numeric(1)))
  correctionFactor <- sqrt( nWithinGroups / (nWithinGroups-1) )
  # Apply the correction factor
  ndatac$sd <- ndatac$sd * correctionFactor
  ndatac$se <- ndatac$se * correctionFactor
  ndatac$ci <- ndatac$ci * correctionFactor
  # Combine the un-normed means with the normed results
  merge(datac, ndatac)
## Norms the data within specified groups in a data frame; it normalizes each
## subject (identified by idvar) so that they have the same mean, within each group
## specified by betweenvars.
##   data: a data frame.
##   idvar: the name of a column that identifies each subject (or matched subjects)
##   measurevar: the name of a column that contains the variable to be summariezed
##   betweenvars: a vector containing names of columns that are between-subjects variables
##   na.rm: a boolean that indicates whether to ignore NA's
normDataWithin <- function(data=NULL, idvar, measurevar, betweenvars=NULL,
                           na.rm=FALSE, .drop=TRUE) {
  require(plyr)
  # Measure var on left, idvar + between vars on right of formula.
  data.subjMean <- ddply(data, c(idvar, betweenvars), .drop=.drop,
                         .fun = function(xx, col, na.rm) {
                           c(subjMean = mean(xx[,col], na.rm=na.rm))
                         },
                         measurevar,
                         na.rm
  # Put the subject means with original data
  data <- merge(data, data.subjMean)
  # Get the normalized data in a new column
  measureNormedVar <- paste(measurevar, "_norm", sep="")
  data[,measureNormedVar] <- data[,measurevar] - data[,"subjMean"] +
    mean(data[,measurevar], na.rm=na.rm)
  # Remove this subject mean column
  data$subjMean <- NULL
  return(data)

```{r print_info_on_design}
cat(paste('number of subjects = ',30,'\n',
          '---','\n',
          'number of trials per subject = ',256,'\n', 
          '---','\n',          
          'number of unique cells in the design = Item(4) * Number(2) * Vagueness(2) * Order(2) * Quantity(2) = ',4*2*2*2*2,'\n',
          'number of times a subject saw a unique cell in the design = 256 / 64 = ', 256/64,'\n',
          '---','\n',
          'number of higher-level cells in the design (abstracting over Order and Quantity) = Item(4) * Number(2) * Vagueness(2) = ',4*2*2,'\n',
          'number of times a subject saw a higher level cell in the design =  = 256 / 16 = ', 4*2*2, '\n',
          sep=''))
```

# Distributional stuff, including transformations.

```{r transformations_table, results='asis'}
x=445
lambda = c(-2,-1,-0.5, 0,0.5,1,2)

xd=data.frame(
  lambda = c(
    lambda[1],
    lambda[2],
    lambda[3],
    lambda[4],
    lambda[5],
    lambda[6],
    lambda[7]
  ),
  x=c(    
    paste("x = ", sep="", x),
    paste("x = ", sep="", x),
    paste("x = ", sep="", x),
    paste("x = ", sep="", x),
    paste("x = ", sep="", x),
    paste("x = ", sep="", x),
    paste("x = ", sep="", x)
     ),
  conventional=c(
    '$$y=\\frac{1}{x^2}$$', 
    '$$y=\\frac{1}{x}$$', 
    '$$y=\\frac{1}{\\sqrt{x}}$$', 
    '$$y=log_e(x)$$', 
    '$$y=\\sqrt{x}$$', 
    '$$y=x$$', 
    '$$y=x^2$$'
    ),
  power=c(
    '$$y=x^{-2}$$', 
    '$$y=x^{-1}$$', 
    '$$y=x^{-0.5}$$', 
    '$$y=log_e(x)$$', 
    '$$y=x^{0.5}$$', 
    '$$y=x^{1}$$', 
    '$$y=x^{2}$$'
    ),
  y=c(     
    paste("y = ", sep="", format(round(x^lambda[1],6), sc=F) ),
    paste("y = ", sep="", round(x^lambda[2],3) ),
    paste("y = ", sep="", round(x^lambda[3],3) ),
    paste("y = ", sep="", round(log(x),3) ),
    paste("y = ", sep="", round(x^lambda[5],2) ),
    paste("y = ", sep="", round(x^lambda[6],6) ),
    paste("y = ", sep="", round(x^lambda[7],6) )
  ),
  back = c(
    '$$x=y^{\\frac{1}{-2}}$$',
    '$$x=y^{\\frac{1}{-1}}$$', 
    '$$x=y^{\\frac{1}{-0.5}}$$', 
    '$$x=exp(y)$$',  
    '$$x=y^{\\frac{1}{0.5}}$$', 
    '$$x=y^{\\frac{1}{1}}$$',  
    '$$x=y^{\\frac{1}{2}}$$'
  ),
  back_lambda=c(
    '$$x=y^{1/\\lambda}$$',
    '$$x=y^{1/\\lambda}$$',
    '$$x=y^{1/\\lambda}$$',
    '$$x=exp(y)$$',  
    '$$x=y^{1/\\lambda}$$',
    '$$x=y^{1/\\lambda}$$',
    '$$x=y^{1/\\lambda}$$'
  ),
  y_back=c(  
    paste( 'x = ', (x^lambda[1]) ^ {1/lambda[1]} ,sep=""),
    paste( 'x = ', (x^lambda[2]) ^ {1/lambda[2]} ,sep=""),
    paste( 'x = ', (x^lambda[3]) ^ {1/lambda[3]} ,sep=""),
    paste( 'x = ', exp(log(x)), sep=""),
    paste( 'x = ', (x^lambda[5]) ^ {1/lambda[5]} ,sep=""),
    paste( 'x = ', (x^lambda[6]) ^ {1/lambda[6]} ,sep=""),
    paste( 'x = ', (x^lambda[7]) ^ {1/lambda[7]} ,sep="")
  )
)

print(xtable(xd, digits=1, align=c(rep('l', times=ncol(xd)+1))), type='html', include.rownames=F, digits=1)
```

Load the file dat.Rda if it exists, else Get the data if file dat.Rda does not already exist.

```{r, get_the_data}
# if the output file dat.Rda already exists then load it, else do data wrangling
if( file.exists('dat.Rda') && file.exists('dd.Rda') ) {
	load('dat.Rda'); load('dd.Rda')

@@

			paste(data_dir,'subject1.data',sep=''),

@@

			current.filename <- paste(data_dir, 'subject', subject, '.data', sep='')

@@

	# save dat as dat.Rda
	# This data set contains *all* trials 7680 including impossible trials and is

@@

	save(dat, file="dat.Rda")

@@

	save(dd, file="dd.Rda")
} # end of the else clause, which is run if both of the files dat.Rda and dd.Rda do not exist.

@@

```{r plotDists, fig.width=20, fig.height=12, message=FALSE}
## plot the comparison of the transformations against their normal equivalents to see distributional properties using full data
# first create a data frame specially for transformations
# this won't be used for analysis, just for this plot
dat_transforms <- dat
# dfSubset says whether any data points were removed
dat_transforms$dfSubset <- "original"
# RTtype says whether the row contains data for an observed RT or 
# a sample from a normal distribution with the same mean and sd
dat_transforms$RTtype <- "observed"
# create extra rows for dfs with removed (outlier) data points
dat1 <- dat_transforms                                       # nrow 7680
dat2 <- subset(dat_transforms, RT>1)                         # nrow 7678
dat2$dfSubset <- "RT>1"
dat3 <- subset(dat_transforms, RT>1 & RT<59999)              # nrow 7677
dat3$dfSubset <- "RT>1&RT<59999"
dat4 <- subset(dat_transforms, RT>1 & RT<59999 & RT<30001)   # nrow 7658
dat4$dfSubset <- "RT>1&RT<30001"
# bring the dfs back into main df
dat_transforms <- rbind(dat1, dat2, dat3, dat4)
dat_transforms$dfSubset <- factor(dat_transforms$dfSubset, levels=c("original", "RT>1", "RT>1&RT<59999", "RT>1&RT<30001"))
# create extra rows for normally distributed transformed variables
dato = dat_transforms
datn = dat_transforms; datn$RTtype <- "normal"
dat_transforms <- rbind(dato,datn)
dat_transforms$RTtype <- factor(dat_transforms$RTtype, levels=c("observed","normal"))
# for safety
dat_transforms[dat_transforms$RTtype=="normal", c("RT_RecSqd", "RT_Rec", "RT_RecSqt", "RT_log", "RT_sqt", "RT_raw", "RT_sqd")] <- NA
# Create the normally distributed transformed variables, replacing the NAs declared above
for (k in levels(dat_transforms$dfSubset)){
  dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="normal", "RT_RecSqd"]  <- 
    rnorm(n=length(dat_transforms[dat_transforms$dfSubset==k  & dat_transforms$RTtype=="observed", "RT_RecSqd"]), 
          mean=mean(dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="observed", "RT_RecSqd"]), 
          sd=sd(dat_transforms[dat_transforms$dfSubset==k     & dat_transforms$RTtype=="observed", "RT_RecSqd"]))
  dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="normal", "RT_Rec"]  <- 
    rnorm(n=length(dat_transforms[dat_transforms$dfSubset==k  & dat_transforms$RTtype=="observed", "RT_Rec"]), 
          mean=mean(dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="observed", "RT_Rec"]), 
          sd=sd(dat_transforms[dat_transforms$dfSubset==k     & dat_transforms$RTtype=="observed", "RT_Rec"]))
  dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="normal", "RT_RecSqt"]  <- 
    rnorm(n=length(dat_transforms[dat_transforms$dfSubset==k  & dat_transforms$RTtype=="observed", "RT_RecSqt"]), 
          mean=mean(dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="observed", "RT_RecSqt"]), 
          sd=sd(dat_transforms[dat_transforms$dfSubset==k     & dat_transforms$RTtype=="observed", "RT_RecSqt"]))
  dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="normal", "RT_log"]  <- 
    rnorm(n=length(dat_transforms[dat_transforms$dfSubset==k  & dat_transforms$RTtype=="observed", "RT_log"]), 
          mean=mean(dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="observed", "RT_log"]), 
          sd=sd(dat_transforms[dat_transforms$dfSubset==k     & dat_transforms$RTtype=="observed", "RT_log"]))
  dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="normal", "RT_sqt"]  <- 
    rnorm(n=length(dat_transforms[dat_transforms$dfSubset==k  & dat_transforms$RTtype=="observed", "RT_sqt"]), 
          mean=mean(dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="observed", "RT_sqt"]), 
          sd=sd(dat_transforms[dat_transforms$dfSubset==k     & dat_transforms$RTtype=="observed", "RT_sqt"]))
  dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="normal", "RT_raw"]  <- 
    rnorm(n=length(dat_transforms[dat_transforms$dfSubset==k  & dat_transforms$RTtype=="observed", "RT_raw"]), 
          mean=mean(dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="observed", "RT_raw"]), 
          sd=sd(dat_transforms[dat_transforms$dfSubset==k     & dat_transforms$RTtype=="observed", "RT_raw"]))
  dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="normal", "RT_sqd"]  <- 
    rnorm(n=length(dat_transforms[dat_transforms$dfSubset==k  & dat_transforms$RTtype=="observed", "RT_sqd"]), 
          mean=mean(dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="observed", "RT_sqd"]), 
          sd=sd(dat_transforms[dat_transforms$dfSubset==k     & dat_transforms$RTtype=="observed", "RT_sqd"]))
} # end of create normally distributed equivalents for the data transforms.
temp <- reshape2::melt(dat_transforms, 
             id.vars=c("dfSubset", "RTtype", "Item", "Vagueness", "Number", "Quantity", "Order"),
             measure.vars=c("RT_RecSqd", "RT_Rec", "RT_RecSqt", "RT_log", "RT_sqt", "RT_raw", "RT_sqd"),
             variable.name="transformation",
             value.name="score")
ggplot(temp, aes(score, colour=RTtype)) + 
  geom_freqpoly() + 
  facet_wrap(dfSubset~transformation, scales="free", nrow=4) +
  theme_bw() + 
	theme(aspect.ratio=1) + 
	scale_color_manual(values=c("blue","red"))

@@

```{r boxPlots, fig.width=20, fig.height=5}
# allow reference to transformation
subdata = reshape2::melt(dd,
               measure.vars=c("RT_Rec", "RT_RecSqt", "RT_log", "RT_raw"),
               variable.name="transformation",
               value.name="value") 
# boxplots for subjects and items separately for each dependent variable
ggplot(subdata) + 
  facet_wrap(~ transformation, scales="free", ncol=4) +
  geom_boxplot(aes(y=value, x=Item, col="Items")) + 
  geom_boxplot(aes(y=value, x=Subject, col="Subjects")) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), aspect.ratio=1, panel.grid=element_blank()) + 
  scale_colour_manual("",values=c("Items"="red","Subjects"="blue")) +
  xlab("") + ylab("")
```

Show how mean times for individual subjects and items vary with respect to the grand mean RT.

```{r fastSlowRT, fig.width=10, fig.height=5}
# Show how mean times for individual subjects and items vary with respect to the grand mean RT.
subs=ddply(dd, .(level=Subject), 
					 summarise, effect="subject", emean=mean(dd$RT), lmean=mean(RT),
					 ldiff=mean(RT)-mean(dd$RT), lsign=ifelse(mean(RT)-mean(dd$RT)>0, "slow", "fast"))
itms=ddply(dd, .(level=Item), 
					 summarise, effect="item", emean=mean(dd$RT), lmean=mean(RT), 
					 ldiff=mean(RT)-mean(dd$RT), lsign=ifelse(mean(RT)-mean(dd$RT)>0, "slow", "fast"))
si=rbind(subs,itms)
si$effect <- factor(si$effect, levels=c("subject", "item"))
si$lsign <- factor(si$lsign, levels=c("slow","fast"))
my1 <- xyplot(data=si, lmean~level | effect, 
              aspect=1,
              layout=c(2,1),
              ylim = extendrange(si$lmean, f = 0.10),
              par.settings=list(trellis.par.set(superpose.symbol = list(col = c("blue","red"), pch=19), 
              																	superpose.line=list(col = c("blue","red")))),
              groups=si$lsign,
              xlab=NULL,
              ylab="RT (in ms)",
              key=list(text=list(levels(si$lsign)), 
              				 lines=list(lty=c(1,1), col=trellis.par.get('superpose.line')$col, 
              				 					 pch=c(19,19), type='o'), divide=1),
              mean=unique(si$emean),
              scales=list(x=list(relation="free", at=list(15.5, 32.5), 
              									 labels=list("subject number", "item identifier"), cex=1 ),
                          y=list(at=c(0, 2000, unique(si$emean), 4000, 6000, 8000,  10000), 
                          			 labels=c(0, 2000,  "Mean", 4000,  6000, 8000, 10000), cex=.9) ),
              panel=function(x,y,type,subscripts,groups=groups,...){
                panel.xyplot(x,y,type='p',groups=groups, subscripts=subscripts,...)
                panel.arrows(x0=x, y0=unique(si$emean), x1=x, y1=y, groups=groups, subscripts=subscripts, 
                						 code=3, length=0, col=trellis.par.get('superpose.line')$col[groups][subscripts], ...)
                panel.abline(h=unique(si$emean), col="lightgrey",lty=1)
                panel.text(x, y=(ifelse(y<unique(si$emean),y-300,y+300)), 
                					 as.numeric(si$level)[subscripts][panel.number()==1], cex=.75)
                panel.text(x, y=(ifelse(y<unique(si$emean),y-300,y+300)), 
                					 as.character(si$level)[subscripts][panel.number()==2], cex=.75)
              }) # end of my1
my2 <- xyplot(data=si, ldiff~level | effect, 
              aspect=1, type='n',
              ylab="Deviation from mean RT (in ms)",
              ylim=extendrange(si$ldiff,f=0.10)
							) # creates a second y-axis scale
print(doubleYScale(my1, my2, 
									 add.axis=TRUE, 
									 add.ylab2=TRUE, 
									 use.style=FALSE))
```

Show how mean times for individual subjects and items vary with respect to the grand mean RT_RecSqt.

```{r fastSlowRecSqt, fig.width=10, fig.height=5}
# Show how mean times for individual subjects and items vary with respect to the grand mean RT_RecSqt.
subs <- ddply(dd, .(level=Subject), summarise, 
              effect="subject", emean=mean(dd$RT_RecSqt), lmean=mean(RT_RecSqt), 
							ldiff=mean(RT_RecSqt)-mean(dd$RT_RecSqt),
              lsign=ifelse(mean(RT_RecSqt)-mean(dd$RT_RecSqt)<0,"slow","fast"))
itms <- ddply(dd, .(level=Item), summarise, effect="item", emean=mean(dd$RT_RecSqt), lmean=mean(RT_RecSqt),
              ldiff=mean(RT_RecSqt)-mean(dd$RT_RecSqt), 
							lsign=ifelse(mean(RT_RecSqt)-mean(dd$RT_RecSqt)<0,"slow","fast"))
si <- rbind(subs,itms)
si$effect <- factor(si$effect, levels=c("subject", "item"))
si$lsign <- factor(si$lsign, levels=c("slow","fast"))
my1 <- xyplot(data=si, lmean~level | effect, 
              aspect=1,
              layout=c(2,1),
              ylim = extendrange(si$lmean, f = 0.10),
              par.settings = list(trellis.par.set(superpose.symbol = list(col = c("blue","red"), pch=19), superpose.line=list(col = c("blue","red")))),
              groups = si$lsign,
              xlab = NULL,
              ylab = "RT_RecSqt: reciprocal of square root of RT in ms\n",
              key=list(text=list(levels(si$lsign)), 
              				 lines=list(lty=c(1,1), 
              				 					 col=trellis.par.get('superpose.line')$col,
              				 					 pch=c(19,19), type='o'),
              				 divide=1),
              mean = unique(si$emean),
              scales = list(x=list(relation="free", at=list(15.5, 32.5), 
              										 labels=list("subject number", "item identifier"), cex=1 ) ),
              panel=function(x,y,type,subscripts,groups=groups,...){
                panel.xyplot(x,y,type='p',groups=groups, subscripts=subscripts,...)
                panel.arrows(x0=x, y0=unique(si$emean), x1=x, y1=y, 
                						 groups=groups, subscripts=subscripts, code=3, length=0, col=trellis.par.get('superpose.line')$col[groups][subscripts], ...)
                panel.abline(h=unique(si$emean), col="lightgrey",lty=1)
                panel.text(x, y=(ifelse(y<unique(si$emean),y-.001,y+.001)), 
                					 as.numeric(si$level)[subscripts][panel.number()==1], cex=.75)
                panel.text(x, y=(ifelse(y<unique(si$emean),y-.001,y+.001)), 
                					 as.character(si$level)[subscripts][panel.number()==2], cex=.75)
              })
my2 <- xyplot(data=si, ldiff~level | effect, 
              aspect=1, type='n',
              ylab="\nDeviation from mean RT_RecSqt (in ms)",
              ylim=extendrange(si$ldiff,f=0.10))
print(doubleYScale(my1, my2, add.axis=TRUE, add.ylab2=TRUE, use.style=FALSE))
```

Show how practice and fatigue effects vary over subjects

```{r pracFatigueSubjs, eval=TRUE, fig.width=20, fig.height=10, cache=FALSE, echo=TRUE}
# Show how practice/fatigue effects vary over subjects
ggplot(dd, aes(y=RT_RecSqt, x=Trial)) + 
	facet_wrap(~Subject,nrow=3) + 
	theme(panel.grid=element_blank(), 
				panel.margin = unit(0, "lines"), 
				panel.background = element_rect(fill = 'white', colour='black' ), 
				legend.position="top") + 
	scale_x_continuous("Trial", breaks=NULL) + 
	geom_point(pch=19, cex=.25, alpha=.5) + 
	stat_smooth(fill='lightblue',col="red",lwd=.5, show.legend=TRUE) + 
	theme(aspect.ratio = 1) 
```

Show practice and fatigue effects vary over items.

```{r pracFatigueItms, eval=TRUE, fig.width=9, fig.height=4, cache=FALSE, echo=TRUE}
# Show how practice/fatigue effects vary over items
ggplot(dd, aes(y=RT_RecSqt, x=Trial)) + 
	facet_grid(~Item) + 
	theme(panel.grid=element_blank(), panel.margin = unit(0, "lines"), 
				panel.background = element_rect(fill = 'white', colour='black' ), 
				legend.position="top") + 
	scale_x_continuous("Trial") + 
	geom_point(pch=19, cex=.25, alpha=.5) + 
	stat_smooth(fill='lightblue',col="red",lwd=.5, show.legend=TRUE) + 
	theme(aspect.ratio = 1) 
```

@@

```{r 4way}
rts <- summarySEwithin(dd, measurevar="RT_RecSqt", withinvars=c("Number", "Vagueness",  "Quantity", "Order"))
rts$condition=paste(rts$Number,rts$Vagueness, sep= ' ')
ggplot(rts, aes(y=RT_RecSqt, x=Vagueness, ymin=RT_RecSqt-ci, ymax=RT_RecSqt+ci, group=Quantity, pch=Quantity)) +
  facet_grid(~Order+Number, scale="free_x") +
  geom_line() +
  geom_errorbar(width=.1) +
  geom_point(cex=3) + 
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.title=element_blank())
```

@@

>>>>

---
title: "Experiment 2 (of 4)"
author: "Matt Green"
output: 
  pdf_document:
    toc: true
---

```{r loadKnitr, echo=FALSE}
suppressPackageStartupMessages(library(knitr))
```

```{r loadLibs, echo=FALSE, message=FALSE}
library(beepr)                     #
library(car)                       #
library(ggplot2)                   #
library(grid)                      #
library(gridExtra)                 #
library(languageR)                 #
library(lattice)                   #
library(latticeExtra)              # for doubleYScale
library(lme4)                      #
library(lmerTest)                  #
library(LMERConvenienceFunctions)  # for perSubjectTrim
library(MASS)                      #
library(plyr)                      #
library(reshape2)                  #
library(rms)                       # for varclus
library(Rmisc)                     # for summarySEwithin and friends
library(xtable)                    #
```

```{r get_the_data, echo=FALSE}
# if the file data.Rda already exists then load it, else do data wrangling
if( file.exists('data_processed.Rda') ) {
	load('data_processed.Rda')

@@

			paste(data_dir,'subject01.data',sep=''),

@@

			current.filename <- paste(data_dir, 'subject', sprintf("%02d",subject), '.data', sep='')

@@

	# This data set (dat) contains *all* trials 7680 including impossible trials and is

@@

	save(dat, file='data_raw.Rda')

@@

	save(dd, file="data_processed.Rda")
} # end of the else clause, which is run if the file data.Rda does not exist.

@@

```{r plotDists, fig.width=5, fig.height=5, message=FALSE, echo=FALSE}
# first create a data frame specially for transformations
# this won't be used for analysis, just for this plot
load('data_raw.Rda') # dat
dat_transforms <- dat
# dfSubset says whether any data points were removed
dat_transforms$dfSubset <- "original"
# RTtype says whether the row contains data for an observed RT or 
# a sample from a normal distribution with the same mean and sd
dat_transforms$RTtype <- "observed"
# create extra rows for dfs with removed (outlier) data points
dat3 <- subset(dat_transforms, RT>1 & RT<59999)              # nrow 7677
dat3$dfSubset <- "RT>1&RT<59999"
dat4 <- subset(dat_transforms, RT>1 & RT<59999 & RT<30001)   # nrow 7658
dat4$dfSubset <- "RT>1&RT<30001"
# bring the dfs back into main df
dat_transforms <- rbind(dat3, dat4)
dat_transforms$dfSubset <- factor(dat_transforms$dfSubset, levels=c("RT>1&RT<59999", "RT>1&RT<30001"))
# create extra rows for normally distributed transformed variables
dato = dat_transforms
datn = dat_transforms; datn$RTtype <- "normal"
dat_transforms <- rbind(dato,datn)
dat_transforms$RTtype <- factor(dat_transforms$RTtype, levels=c("observed","normal"))
# for safety
dat_transforms[dat_transforms$RTtype=="normal", c("RT_log", "RT_raw")] <- NA
# Create the normally distributed transformed variables, replacing the NAs declared above
for (k in levels(dat_transforms$dfSubset)){
    dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="normal", "RT_log"]  <- 
    rnorm(n=length(dat_transforms[dat_transforms$dfSubset==k  & dat_transforms$RTtype=="observed", "RT_log"]), 
          mean=mean(dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="observed", "RT_log"]), 
          sd=sd(dat_transforms[dat_transforms$dfSubset==k     & dat_transforms$RTtype=="observed", "RT_log"]))
    dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="normal", "RT_raw"]  <- 
    rnorm(n=length(dat_transforms[dat_transforms$dfSubset==k  & dat_transforms$RTtype=="observed", "RT_raw"]), 
          mean=mean(dat_transforms[dat_transforms$dfSubset==k & dat_transforms$RTtype=="observed", "RT_raw"]), 
          sd=sd(dat_transforms[dat_transforms$dfSubset==k     & dat_transforms$RTtype=="observed", "RT_raw"]))
  }# end of create normally distributed equivalents for the data transforms.
temp <- reshape2::melt(dat_transforms, 
             id.vars=c("dfSubset", "RTtype", "Item", "Vagueness", "Number", "Quantity", "Order"),
             measure.vars=c("RT_log", "RT_raw"),
             variable.name="transformation",
             value.name="time")
ggplot(temp, aes(time, colour=RTtype)) + geom_freqpoly() + facet_wrap(transformation~dfSubset, scales='free') + scale_color_manual(values=c("blue","red")) + theme_bw() + theme(aspect.ratio=1)
```

@@

```{r boxPlots, fig.width=7, fig.height=3, echo=FALSE}
load('data_processed.Rda')
# allow reference to transformation
subdata = reshape2::melt(dd,
               measure.vars=c("RT_log", "RT_raw"),
               variable.name="transformation",
               value.name="value") 
# boxplots for subjects and items separately for each dependent variable
ggplot(subdata) + 
  facet_wrap(~ transformation, scales="free", ncol=4) +
  geom_boxplot(aes(y=value, x=Item, col="Items"), outlier.colour = "black", outlier.shape = 21, outlier.size=0.1) + 
  geom_boxplot(aes(y=value, x=Subject, col="Subjects"), outlier.colour = "black", outlier.shape = 21, outlier.size=0.1) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5), aspect.ratio=1, panel.grid=element_blank()) + 
  scale_colour_manual("",values=c("Items"="orange","Subjects"="blue")) +
  xlab("") + ylab("")
```

Show how mean times for individual subjects and items vary with respect to the grand mean Log RT.

```{r fastSlowRT, fig.width=7, fig.height=3, message=FALSE, echo=FALSE}
# Show how mean times for individual subjects and items vary with respect to the grand mean RT.
subs=ddply(dd, .(level=Subject), 
					 summarise, effect="subject", emean=mean(dd$RT_log), lmean=mean(RT_log),
					 ldiff=mean(RT_log)-mean(dd$RT_log))
itms=ddply(dd, .(level=Item), 
					 summarise, effect="item", emean=mean(dd$RT_log), lmean=mean(RT_log), 
					 ldiff=mean(RT_log)-mean(dd$RT_log))
si=rbind(subs,itms)
si$effect <- factor(si$effect, levels=c("subject", "item"))
my1 <- xyplot(data=si, lmean~level | effect, 
              aspect=1,
              layout=c(2,1),
              ylim = extendrange(si$lmean, f = 0.10),
              par.settings=list(trellis.par.set(superpose.symbol = list(col='black'), 
              																	superpose.line=list(col='black'))),
              xlab=NULL,
              ylab=NULL,
              main='How mean times for individual subjects and items\nvary with respect to the grand mean Log RT',
              mean=unique(si$emean),
              scales=list(x=list(relation="free", at=NULL, 
              									 labels=list("subject number", "item identifier"), cex=1 ),
                          y=list(at=    c(0,7,unique(si$emean),8,9),
                          			 labels=c("0","7","Mean Log RT","8","9"))),
              panel=function(x,y,type,subscripts,...){
                panel.abline(h=unique(si$emean), col="lightgrey",lty=1,...)
                panel.arrows(x0=x, y0=unique(si$emean), x1=x, y1=y, subscripts=subscripts, code=3, length=0, ...)
                panel.xyplot(x,y,type='p', pch=21, cex=.5, col='black', fill='black', subscripts=subscripts,...)
              }) # end of my1
my2 <- xyplot(data=si, ldiff~level | effect, 
              aspect=1, type='n',
              ylab="Deviation from\nmean Log RT",
              ylim=extendrange(si$ldiff,f=0.10)
							) # creates a second y-axis scale
print(doubleYScale(my1, my2, 
									 add.axis=TRUE, 
									 add.ylab2=TRUE, 
									 use.style=FALSE))
```

@@

```{r 4way, fig.width=6, fig.height=3, echo=FALSE}
rts <- summarySEwithin(dd, measurevar="RT_RecSqt", withinvars=c("Number", "Vagueness",  "Quantity", "Order"))
rts$condition=paste(rts$Number,rts$Vagueness, sep= ' ')
ggplot(rts, aes(y=RT_RecSqt, x=Vagueness, ymin=RT_RecSqt-ci, ymax=RT_RecSqt+ci, group=Quantity, pch=Quantity)) +
  facet_grid(~Order+Number, scale="free_x") +
  geom_line() +
  geom_errorbar(width=.1) +
  geom_point(cex=3) + 
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.title=element_blank())
```

@@

```{r v511, results='asis'}
print(xtable(coef(summary(v5b))))
```
